---
title: "MTH6139 Time Series" 
subtitle: "Deaths England and Wales (Jan 2020 - Mar 2025)" 
author: "Oscar John George Hope" 
date: "Spring term 2025" 
output: 
  html_document:
    toc: true
    toc_float: true
    theme: spacelab 
    highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE}
# This code will display the QMUL logo at the top right of the page
# Do not change this code
htmltools::img(src = knitr::image_uri("images/QMlogo.png"),
               alt = 'logo',
               style = 'position:absolute; top:0; right:0; padding:10px; width:20%;')
```

# Introduction

## 1.1 The times series data of deaths in England and Wales (Jan 2020 - Mar 2025, weekly)

```{r}
death_statistics
```

The data above shows the deaths in England and Wales over the period
from January 2020 to March 2025 in weekly intervals, the date on the ds
axis is the final day of the week in measure.

## 1.2 The purpose of this project.

I have decided to undertake this project because my family have long
time been in the funeral industry and as a result I have been doing lots
of data interpretation with funerals, deaths and and hospitals since I
came to university and in other statistical modules to get a better
understanding for my parents business and potentially help, as far as I
am aware they do not do any kind of modelling like this.

## 1.3 Graph plots and tables

### 1.3.1 forecasting

```{r}
plot(m,p)
```

```{r}
print(forecast_table)
```

Graph:

The graph above shows the historic data taken from the imported death
statistics and an additional forecast for 28 periods after that in
months. the graph in the historical part seems to be repeating annually
with the same effect meaning one cycle is a year. From this we can see
it could be described as a spiking upward sine function.

I obtained this graph by doing the following code:\
First i intalled the correct packages being 'remotes' which helps me
install packages from 'GitHub through which i installed 'Prophet'
package, the i loaded libraries i may need (Rcpp, prophet(TS
forecasting), astsa (statistictial analysis not used here) and
readxl(for reading my excel file)). Next i import my data
'death_statistics.xlsx' using the 'readxl'. this i move to forecasting
using 'prophet' to fit my data where i had to specify my data was weekly
not daily, i stored this as 'm'. I then forcast the data in the future
using 'make_future_dataframe' for 28 periods with frequency of months, I
store this as 'f'. I can then use 'predict' for 'm' and 'f' to finally
plot the predicted in 'p' against me TS model 'm'.

Table:

I wanted to see the values for the predicted values in the the graph as
it is difficult to see. maily to see exactly what months these spikes
occur as i could also see a small spike near the start of the year. From
this table i could see that small spike in in the 4th month out of the
yeah likely being near the 23rd due to the symmetry of the 3rd and 5th
month. i also found the trough is in 7th and 8th month likely around the
9th of the 8th month through symmetry again.

I produced the table though only considering the future values by taking
the dates after the most recent in the origional data using

```{r}
forecast_data <- p[p$ds >= max(death_statistics$ds), ]

```

Then I can take the relevant data from the columns of the forecast such
as 'yhat_lower' 'yhat_upper' and 'yhat' and put it into a table before
naming the columns.

### 1.3.2 Data decomposition

```{r}
plot(decompose(ts_data_for_decompose))
```

Above is my decomposition of data for the period we can see the trend
there is a spike at the start probably due to COVID but he seasonality
is very similar to what we observed from our previous table meaning this
data is reliant alone on its seasonal aspect not so much its trend later
on after 2021. The seasonality as what might be expected suggest more
deaths in the winter at the end of the year perhaps from illness like
influenza and other viruses that are more prominent in the colder
months.

I created this by assigning values to a time series for frequency of 52
due to it being in weeks starting at the start of 2020. and then
plotting using the decompose of the time series into the plot.

```{r}
plot(dd.df$t, dd.df$x, type='l',xlab='Time',ylab='Deaths',main='Weekly Death Statistics')
model <-lm(x~t,data=dd.df)
lines(dd.df$t,fitted(model),col='green')
```

The graph above shows a plot of observations from the data frame against
time in the data frame. I then put onto the line of linear regression
for the time which i don't believe is very helpful due to the cyclical
nature of the data. The line of linear regression is going down but very
shallow gradient. I don't believe wide spread immortality in humans is
very soon, this downward nature is likely due to the spike in deaths
during COVID.

I coded this graph by simply plotting from the data frame using the code
in the R box above labelling as well. then making a linear model from
the data from the data frame and fitting that as a line over the top of
the other graph in green to keep it light hearted and not morbid.

### 1.3.3 Other time frames

```{r}
f =prophet::make_future_dataframe(m, periods=1000, freq="day")
p =predict(m, f)
plot(m,p)
```

```{r}
f =prophet::make_future_dataframe(m, periods=12, freq="quarter")
p =predict(m, f)
plot(m,p)
```

Above i have used two other time frames, day and quarter. The day
requires me to have many periods, in this example 1000 which would can
take a lot longer to load and it unnecessary to have make it a day due
to the randomness of deaths in a day period and autocorrelation of the
percent movements in day to day date being low I would consider this
overkill especially as the original data is in weeks.\
The other being every quarter does not give enough information, the
users of this information will not be looking for 100 years into the
future only likely 5 years making periods 20 is well within the power of
R studios. also the graph does not look particularly helpful in
identifying the individual peeks in January and April.

# References

-   Datacamp course:
    <https://www.datacamp.com/courses/reporting-with-rmarkdown>
-   RStudio reference: <https://rmarkdown.rstudio.com/lesson-1.html>
-   More sophisticated reference:
    <https://bookdown.org/yihui/rmarkdown/>
-   Quick Start \| Prophet:
    <https://www.statista.com/statistics/1111804/weekly-deaths-in-england-and-wales/>
-   ts function - RDocumentation: <https://www.rdocumentation.org/>
-   Forecasting: Principles and Practice (2nd ed):
    <https://otexts.com/fpp2/tspatterns.html>
